Mergeable Types

### 2017-02-28 paper/mergeable-types.tex
edit

\noindent Rather than specifying merge ($M$) in terms of an LCA as we
did in describing our programming model,
Def.~\ref{def:mergeable-type} specifies $M$ in terms of a pair
of morphisms.  This specification is useful to avoid the expression of
certain nonsensical merge functions, and thus enforce a useful
branch-local consistency property
(Theorem~\ref{thm:branch-consistency}). For instance, consider the
counter data type from Sec.~\ref{sec:intro}. A three-way merge
function can be defined to return a constant regardless of its
arguments:
\begin{ocaml}
  let merge lca a b = 0
\end{ocaml}
Such a merge function allows the client to witness a violation of
branch-local consistency. For example, the client may
\C{sync} the counter, whose current local value is $10$, and obtain
$0$ as a result, thus witnessing a violation of a counter's monotonicity
property. Fortunately, Def.~\ref{def:mergeable-type} disallows this
semantics. It is impossible to define the function $M$ for the counter
data type that violates the monotonicity invariant, because no counter
morphism violates the invariant. In general, the specification of $M$
guarantees that the merge operation for a type $t$ preserves any
invariants (e.g., balancedness of a tree, sortedness of a list,
non-negativeness of an integer etc) that are preserved by all of $t$'s
morphisms.  The commutativity of the diagram in
Fig.~\ref{fig:pushouts} need not be enforced statically. It can
equally be verified at run-time by checking that the morphisms
returned by $M$ map respective concurrent versions to the same
value. The \rulelabel{E-Pull-Wait} rule of the operational semantics
(Fig.~\ref{fig:opsem}) can be extended to this effect:
\begin{smathpar}
\begin{array}{c}
\RULE
{
  t\neq t' \spc
  \under{H}{v' \mbleto v} \spc
% \C{world}(H,t') \semsucceq \C{world}(H,t)\spc
  v' \not\preceq v \spc
  (P',Q') = M(\C{lca}(H(t),H(t')), v, v') \spc
  Q'(v) = P'(v')
}
{
  (\pull)_t;H(t \mapsto (v,f)::m)(t' \mapsto (v',\_)::\_) ~\stepsto~
  (\pull)_t;H[t \mapsto (v_m,\C{MERGE}\; H(t'))::(v,f)::m]
}
\end{array}
\end{smathpar}

>>>>

Rather than specifying merge ($M$) in terms of an LCA as we
did in describing our programming model,
Def.~\ref{def:mergeable-type} specifies $M$ in terms of a pair
of morphisms.  This specification is useful to avoid the expression of
certain nonsensical merge functions, and thus enforce a useful
branch-local consistency property
(Theorem~\ref{thm:branch-consistency}). For instance, consider the
counter data type from Sec.~\ref{sec:intro}. A three-way merge
function can be defined to return a constant regardless of its
arguments:
\begin{ocaml}
  let merge lca a b = 0
\end{ocaml}
Such a merge function allows the client to witness a violation of
branch-local consistency. For example, the client may
\C{sync} the counter, whose current local value is $10$, and obtain
$0$ as a result, thus witnessing a violation of a counter's monotonicity
property. Fortunately, Def.~\ref{def:mergeable-type} disallows this
semantics. It is impossible to define the function $M$ for the counter
data type that violates the monotonicity invariant, because no counter
morphism violates the invariant. In general, the specification of $M$
guarantees that the merge operation for a type $t$ preserves any
invariants (e.g., balancedness of a tree, sortedness of a list,
non-negativeness of an integer etc) that are preserved by all of $t$'s
morphisms.  The commutativity of the diagram in
Fig.~\ref{fig:pushouts} need not be enforced statically. It can
equally be verified at run-time by checking that the morphisms
returned by $M$ map respective concurrent versions to the same
value. The \rulelabel{E-Pull-Wait} rule of the operational semantics
(Fig.~\ref{fig:opsem}) can be extended to this effect:
\begin{smathpar}
\begin{array}{c}
\RULE
{
  t\neq t' \spc
  \under{H}{v' \mbleto v} \spc
% \C{world}(H,t') \semsucceq \C{world}(H,t)\spc
  v' \not\preceq v \spc
  (P',Q') = M(\C{lca}(H(t),H(t')), v, v') \spc
  Q'(v) = P'(v')
}
{
  (\pull)_t;H(t \mapsto (v,f)::m)(t' \mapsto (v',\_)::\_) ~\stepsto~
  (\pull)_t;H[t \mapsto (v_m,\C{MERGE}\; H(t'))::(v,f)::m]
}
\end{array}
\end{smathpar}
